Inferring the Recombinant Genealogy of a Pandemic Pathogen in Real Time

### 2025-09-19 latex/main.tex
Updates from Overleaf

Ancestral recombination graphs (ARGs) describe the reticulate 
genealogies of sampled sequences in a recombining species\cite{Wong2024}.
While ARGs have been of theoretical interest in population genetics 
for several decades\cite{Hudson1983,Griffiths1991two,Griffiths1997ancestral},
recent breakthroughs in simulation\cite{Kelleher2016,Kelleher2018efficient}
and inference~\citep{Rasmussen2014,Speidel2019,Kelleher2019,
Wohns2022,Zhang2023,Gunnarsson2024,Deng2024robust} have led to a surge 
of interest in their application\cite{Brandt2024,Lewanski2024,Nielsen2025}.

% TODO merge this in
Ancestral Recombination Graphs (ARGs) are reticulate genealogies
that incorporate both mutation and recombination events \cite{Wong2024},
potentially allowing more comprehensive evolutionary representations of SARS-CoV-2
and other viruses undergoing recombination.
Until recently, ARGs lacked software support and
sufficiently scalable inference methods to be of practical use.
Recent advances enable ARG inference for tens to hundreds of thousands of human genomes \cite{Kelleher2019,Speidel2019,Zhang2023},
with high levels of recombination.
These advances now enable evolutionary genealogies to be built quickly and at scale, and
there is a surge of excitement in the inference and applications of ARGs \cite{Brandt2024,Lewanski2024,Wong2024,Nielsen2025}.

@@

\subsubsection*{Ancestral Recombination Graphs}
\label{sec:args}

describe how a set of sample sequences
relate to each other at each position along the genome in a recombining
species 
lots of interest

, and are currently the subject of intense research interest 
(Brandt et al. 2024, Lewanski et al. 2024, Wong et al. 2024, Nielsen et al. 2025).

%%% This paragraph needs to be rewritten.
Originally introduced (Griffiths, 1991; Griffiths and Marjoram, 1997)
to define an alternative formulation of the coalescent with recombination (Hudson, 1983),
the term ``Ancestral Recombination Graph'' has come to be used
more generally to describe not just realizations of the model,
but any recombinant genetic ancestry (Minichiello and Durbin, 2006).
While there is some subtlety in the details \cite{Wong2024},
we can think of an ARG as being any graph
that encodes the reticulate genetic genealogy of a sample of sequences
under the influence of recombination.

Until recently, ARGs lacked software support and
sufficiently scalable inference methods to be of practical use.
Recent advances enable ARG inference for tens to hundreds of thousands of human genomes \cite{Kelleher2019,Speidel2019,Zhang2023},
with high levels of recombination.
These advances now enable evolutionary genealogies to be built quickly and at scale, and
there is a surge of excitement in the inference and applications of ARGs \cite{Brandt2024,Lewanski2024,Wong2024,Nielsen2025}.

@@

We devised a quality control (QC) measure based on this observation, and
applied it to each putative recombination event in the ARG.
We examined the genomic positions at which the suggested parents differed in allelic state.
Where such sites clustered closely (3 or fewer bases apart),
we treated the entire cluster of sites as a single \emph{locus}.
For each genomic region identified by the HMM as having a single parent,
we calculated the "net number of supporting loci" by counting the number of loci in the region
that supported the suggested parent minus the number that did not support it.
Finally, with all the recombination events in the ARG having a single breakpoint,
we calculated the net number of supporting loci on the left and right sides of each breakpoint, and
used this to plot the 855 recombination events in Figure~\ref{fig:recombinant_qc}.

This simple QC measure effectively differentiates
artefactual recombination events from those that are well supported.
All the sc2ts-detected recombination events associated with the Pango X lineages (Table~\ref{tab:pango_x_lineages})
occur in the upper right of Figure~\ref{fig:recombinant_qc},
with high numbers of net supporting loci on both the right and left parents.
In contrast, the other regions of the plot, where one or both parents are less well supported,
are dominated by recombination events associated with samples sequenced using a specific protocol (AmpliSeq).
Coupled with closer inspection of copying patterns (see below),
we chose a QC cutoff of four or more net supporting loci on both the left and right side of the breakpoint.
This leads to 354 QC-passing recombination events (Figure~\ref{fig:recombinant_qc}, upper right quadrant),
compared to 501 recombination events
for which there is less support and which may be artefactual (shaded quadrants).

Inspection of the copying patterns for all the recombination events
clearly identified associations between recombination events that failed QC and
known problematic loci in the SARS-CoV-2 genome.
Figure~\ref{fig:copying_pattern_examples} shows illustrative copying patterns
(Q1 to Q4 correspond to the labeled quadrants in Figure~\ref{fig:recombinant_qc}),
in which adjacent sites are underlined in red, and near-adjacent sites (within 3 bases) in orange.
A continuous red+orange underline thus indicates a region treated as a single locus.
Q1 shows a well supported recombination event that passed QC (albeit marginally),
having a net number of supporting loci of 4 on both sides of the breakpoint.
Q2 shows an artefactual recombination event caused by a 9-base Delta lineage-defining deletion.
There are XX (XX\%) of the QC-failing recombination events involving this deletion, and
many cases are affected by a known amplicon dropout caused by this particular deletion \cite{Sanderson2021}
Q3 shows an artefactual recombination event caused by
a combination of this Delta deletion and a known MNS (see below).
Finally, Q4 shows an artifactual recombination event associated with the A507T-T508C-G509A MNS,
which has been identified as a sequencing error due to incomplete read trimming \cite{DeMaio2024}.
There are XX recombination events involving this particular MNS, of which XX\% failed QC.

>>>>

Ancestral Recombination Graphs (ARGs) are reticulate genealogies
that incorporate both mutation and recombination events \cite{Wong2024},
potentially allowing more comprehensive evolutionary representations of SARS-CoV-2
and other viruses undergoing recombination.
Until recently, ARGs lacked software support and
sufficiently scalable inference methods to be of practical use.
Recent advances enable ARG inference for tens to hundreds of thousands of human genomes \cite{Kelleher2019,Speidel2019,Zhang2023},
with high levels of recombination.
These advances now enable evolutionary genealogies to be built quickly and at scale, and
there is a surge of excitement in the inference and applications of ARGs \cite{Brandt2024,Lewanski2024,Wong2024,Nielsen2025}.

@@

\subsubsection*{Ancestral Recombination Graphs}
\label{sec:args}
%%% This paragraph needs to be rewritten.
Originally introduced (Griffiths, 1991; Griffiths and Marjoram, 1997)
to define an alternative formulation of the coalescent with recombination (Hudson, 1983),
the term ``Ancestral Recombination Graph'' has come to be used
more generally to describe not just realizations of the model,
but any recombinant genetic ancestry (Minichiello and Durbin, 2006).
While there is some subtlety in the details \cite{Wong2024},
we can think of an ARG as being any graph
that encodes the reticulate genetic genealogy of a sample of sequences
under the influence of recombination.

@@

We devised a quality control (QC) measure based on this observation, and
applied it to each putative recombination event in the ARG.
We examined the genomic positions
at which the suggested parents differed in allelic state.
Where such sites clustered closely (3 or fewer bases apart),
we treated the entire cluster of sites as a single \emph{locus}.
For each genomic region identified by the HMM as having a single parent,
we calculated the "net number of supporting loci"
by counting the number of loci in the region
that supported the suggested parent minus the number that did not support it.
Finally, with all the recombination events in the ARG having a single breakpoint,
we calculated the net number of supporting loci
on the left and right sides of each breakpoint, and
used this to plot all the 855 recombination events in Figure~\ref{fig:recombinant_qc}.

This simple QC measure effectively differentiates
artefactual recombination events from those that are well supported.
All the sc2ts-detected recombination events
associated with the Pango X lineages (Table~\ref{tab:pango_x_lineages})
occur in the upper right of Figure~\ref{fig:recombinant_qc},
with high numbers of net supporting loci on both the right and left parents.
In contrast, the other regions of the plot,
where one or both parents are less well supported,
are dominated by recombination events associated with
samples sequenced using the Ion AmpliSeq protocol.
Coupled with closer inspection of copying patterns (see below),
we chose a QC cutoff of 4 or more net supporting loci
on both the left and right side of the breakpoint.
This leads to 354 QC-passing recombination events
(Figure~\ref{fig:recombinant_qc}, upper right quadrant),
compared to 501 recombination events
for which there is less support and which may be artefactual (shaded quadrants).

Inspection of the copying patterns for all the recombination events
clearly identified associations between recombination events that failed QC and
known problematic loci in the SARS-CoV-2 genome.
Figure~\ref{fig:copying_pattern_examples} shows illustrative copying patterns
(Q1 to Q4 correspond to the labeled quadrants in Figure~\ref{fig:recombinant_qc}),
in which adjacent sites are underlined in red, and
near-adjacent sites (within 3 bases) in orange.
A continuous red+orange underline thus indicates a region treated as a single locus.
Q1 shows a well supported recombination event that passed QC (albeit marginally),
having a net number of supporting loci of 4 on both sides of the breakpoint.
Q2 shows an artefactual recombination event
caused by a 6-base Delta lineage-defining deletion.
There are 71 (14\%) QC-failing recombination events involving this deletion, and
many cases are affected by a known amplicon dropout
caused by this particular deletion \cite{Sanderson2021}
Q3 shows an artefactual recombination event caused by
a combination of this Delta deletion and a known MNS (see below).
Finally, Q4 shows an artifactual recombination event
associated with the A507T-T508C-G509A MNS,
which has been identified as a sequencing error
due to incomplete read trimming \cite{DeMaio2024}.
There are 9 recombination events involving this particular MNS, of which 6 failed QC.
